---
title: "Homework 3 - dplyr"
author: "Adam Shah (2369712)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(babynames)
library(tidyverse)
library(nycflights13)
library(plotly)
```

**About the datasets**:

**`babynames`**: Baby names. Full baby name data provided by the SSA. This includes all names with at least 5 uses. The data frame has five variables: year, sex, name, n and prop (n divided by total number of applicants in that year, which means proportions are of people of that gender with that name born in that year).

**`flights`**: Dataset with flights that departed from NYC (JFL, LGA, or EWR airports). To see the description of the variables type `help(flights)`.

**Q1** (2 points): Calculate the total number of children born each year over time, and then make a line graph to show the trend, and set color to the variable sex.

```{r}
names_by_year <- babynames %>% group_by(sex) %>% count(year, sort=T)
names_by_yearplot <- ggplot(names_by_year, aes(year, n, color=sex)) +
  geom_line()
ggplotly(names_by_yearplot)
```

**Q2** (2 points): Identify the top 5 male and female names and show on the barplot, set color for different sex.

```{r}
namecounts <- babynames %>% group_by(name, sex) %>%
  summarize(count = as.numeric(sum(n)))
namesrank <- namecounts %>% group_by(name, sex) %>%
  summarise(count=sum(count)) %>%
  group_by(sex) %>%
  mutate(rank = min_rank(desc(count))) %>%
  filter(rank < 6) %>%
  arrange(sex, rank) %>%
  select(name, sex, count)
print(namesrank)
plot_ly(data=namesrank, x=~name, y=~count, color=~sex, type = 'bar')
```

**Q3** (3 points): For each airport (dest), calculate the total number of flights, the average distance of all flights, and the average arrival delay of all flights, save the results in a dataset (give a name), and remove the airport "HNL" and the airports with less than 20 flights.

```{r}
totalflights <- flights %>% group_by(dest) %>% count(dest)
flightavgs <- flights %>% group_by(dest) %>%
  summarise(avg_distance=mean(distance, na.rm = T), avg_arr_delay=mean(arr_delay, na.rm = T)) 
flightdata <- merge(totalflights, flightavgs, by="dest")
flightdata <- flightdata %>%
  filter(!dest=="HNL") %>% 
  filter(!n < 20)
head(flightdata)
```


**Q4** (3 points): Use the dataset you created, provide a scatterplot to show the relationship between travel distance and arrival delay, set size to the total number of flights and the transparency of 0.5, and add a smooth line without showing the SE.

What information can you get from the graph?

```{r}
flightplot<- ggplot(flightdata, aes(x=avg_distance, y=avg_arr_delay, size=n))+ geom_point(alpha=.5)+ geom_smooth(method="lm", se=F)
flightplot
```

From this graph, I can tell that there is a negative correlation between these two variables- as flight distance increases, the average arrival delay decreases. 