---
title: "DataFest22workshop2"
author: "Zhiyun Gong"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nycflights13)
library(babynames)
library(stringr)
```


##  dplyr package: data manipulation

*split-apply-combine*: A common analytical pattern is to split data frame into pieces, apply some function to each pieces, and combine the results back together again to a new data frame. 

It provides simple *verbs*, functions that correspond to the most common data manipulation tasks, to help you translate your thoughts into code.

*dplyr* aims to provide a function for each basic verb of data manipulation:

- group_by(): to split the data frame into pieces based on some variables.
- filter(): to select cases based on their values.
- arrange(): to give cases/observations a specific order
- select(): to select variables based on their names.
- mutate(): to add new variables that are functions of existing variables.
- summarise(): to condense multiple values to a single value.

All verbs work similarly:
1. The first argument is a data frame;
2. The subsequent arguments describe what to do with the data frame, using the variable names (without quotes);
3. The result is a new data frame;

### Import dataset

```{r}
babynames
```


### filter()
filter() to select cases/observations/rows based on their values.
```{r}
filter(babynames, year==2000)
```




### arrange()
Reorder the observations by their values of a variable
```{r}
# arrange(dataset, variable)
arrange(babynames, n) # ascending order is default

```


*filter() and arrange() work on **observations/rows** *



### select()
select() to select variables based on their names.

```{r}
select(babynames,name, n)

```


### mutate()
create new columns that are functions of existing columns and add them to the original dataset as a new dataset. 
```{r}
mutate(babynames, rank=(min_rank(desc(n)))) # keep all variables including the new one

```

*select() and mutate() work on **variables/columns** *

### Pipe operator: *%>%* (*then*)
Pipes take the output from one function and feed it to the first argument of the next function.

### Pair summarise with group_by(). 
group_by() to split the data frame into pieces based on some variables. 
*split-apply-combine*: split data frame into groups, apply some function to each group, and combine the results back together again to a new data frame. 
```{r}
babynames %>%
  group_by(name) %>%
  summarise(total=sum(n)) %>%
  arrange(desc(total)) %>%
  top_n(5) %>%
  ggplot(aes(x=name, y=total)) + geom_col() + coord_flip()
```




## forcats

This package provides some tools for working with categorical variables (factors)

Question 1: find the relation between religions and average hours of watching tv.

```{r}

relig_summary <- gss_cat %>%
  group_by(relig) %>%
  summarise(tvhours=mean(tvhours,na.rm = T)) 

  ggplot(relig_summary,aes(x=tvhours, y=relig)) +
  geom_point()
```

How could we interpret the plot without overall pattern? 


### fct_reorder()

Reordering the levels of a factor using *fct_reorder()*.

*fct_reorder()* takes three arguments:

- f, the factor whose levels you want to modify.

- x, a numeric vector that you want to use to reorder the levels.

- Optionally, fun, a function used if there are multiple values of x for each value of f. The default value is median.

```{r}
ggplot(relig_summary, aes(x=tvhours, y=fct_reorder(relig, tvhours))) +
  geom_point()

```


### fct_recode


```{r}
levels(gss_cat$partyid)

gssnew<-gss_cat %>%
drop_na(tvhours) %>%
mutate(partyidnew = fct_recode(partyid,
    "Republican"    = "Strong republican",
    "Republican"      = "Not str republican",
    "Independent" = "Ind,near rep",
    "Independent" = "Ind,near dem",
    "Democrat"        = "Not str democrat",
    "Democrat"      = "Strong democrat",
    "Other"                 = "No answer",
    "Other"                 = "Don't know",
    "Other"                 = "Other party")) 
    
group_by(gssnew, partyidnew) %>%
summarise(meantvhours = mean(tvhours)) %>%
ggplot(aes(x=meantvhours, y=fct_reorder(partyidnew,meantvhours))) +
geom_col()

```

### fct_collapse()

If you want to collapse a lot of levels, fct_collapse() is a useful variant of fct_recode(). For each new variable, you can provide a vector of old levels:

```{r}
gss_cat %>%
  mutate(partyidnew = fct_collapse(partyid,
    other = c("No answer", "Don't know", "Other party"),
    rep = c("Strong republican", "Not str republican"),
    ind = c("Ind,near rep", "Independent", "Ind,near dem"),
    dem = c("Not str democrat", "Strong democrat")
  )) %>%
  count(partyidnew)
```



## Joining relational datasets

```{r}

# Toy datasets to use


band <- tribble(
   ~name,     ~band,
  "Mick",  "Stones",
  "John", "Beatles",
  "Paul", "Beatles"
)

instrument <- tribble(
    ~name,   ~plays,
   "John", "guitar",
   "Paul",   "bass",
  "Keith", "guitar"
)

instrument2 <- tribble(
    ~artist,   ~plays,
   "John", "guitar",
   "Paul",   "bass",
  "Keith", "guitar"
)
```

### Mutating joins

#### 1.  *left_join(x, y, by="key variable")*: keeps all the observations (rows) in x

A key is a variable that uniquely identifies an observation, otherwise, we need multiple variables to identify an observation,

```{r}
## left_join(x, y, by="key variable")  
## join y to x, x is the primary dataset
## mutate new variables in x by copying different variables in y
## return a new dataset

## left_join(x, y, by=c("key variable1", "key variable2"))

dat1<-left_join(band, instrument, by="name")
dat1
```

#### 2. *right_join(x, y, by="")*: keep all the observations (rows) in y

The opposite way of *left_join()*

```{r}
right_join(band,instrument, by="name")
```



#### 3. *inner_join()* keeps all the observations in **both** x and y

An inner join keeps observations that appear in both tables. 

```{r}
inner_join(band, instrument, by="name")
#will lose some information in the primary dataset
```


#### 4. *full join* keeps all observations in x and y

An full join keeps observations that appear in either x or y.

```{r}
full_join(band, instrument, by="name")
#better, won't lose primary dataset info
```


### Filtering joins

#### 1. *semi_join(x,y, by="")* keeps all the observations of x that have a match in y


use semi_join() to collect the artists in *band* that have instrument info in *instrument*.
```{r}
semi_join(band, instrument, by="name")

## use semi_join() to collect the artists in *band* that have instrument info in *instrument*.
```



#### 2. *anti_join(x,y, by="")* drops all the observations of x that have a match in y.  

```{r}
anti_join(band, instrument, by="name")

## Use an anti_join() to return the rows of artists for which you don't have any instrument info.
```



#### Example: join relational datasets connected by key variables 

```{r,eval=FALSE}
# check out the package "nycflights13"

flights # connects to 
planes # via a single variable *tailnum*

flights # connects to 
airlines # through the *carrier* variable 

flights # connects to 
airports # via two variables *faa* and *dest*

flights # connects to 
weather # via *origin*, *year*, *month*, *day*, and *hour*

```




Example 1:


Join airports to flights and get how many flights in flights flew to the airports listed in airports?
(Hint: use n distinct() function to count.)

```{r}
flights %>% 
semi_join(airports, by=c("dest"="faa")) %>%  #dest and faa are same thing in different datasets
  summarise(total=n_distinct(dest))

flights %>% 
anti_join(airports, by=c("dest"="faa")) %>%  
  summarise(total=n_distinct(dest))
#number of airports in airports not in flights

```


## tidy up datasets

The *tidyr* package helps tidy up messy datasets. 

There are three interrelated rules which make a dataset tidy:

- Each variable must have its own column

- Each observation must have its own row

- Each value must have its own cell

There are a few key functions in the *tidyr* package, *pivot_longer()*, *pivot_wider()*, *separate()*


```{r,echo=FALSE}
# toy datasets

# population table
table1<-tibble(
  `country`=c("Afghanistan","Brazil","China"),
  `1999`=c(19987071, 172006362, 1272915272),
  `2000`=c(20595360, 174504898, 1280428583)
)
table1
```



### 1. pivot_longer()

A common problem is a dataset where some of the column names are not names of variables, but *values* of a variable.


To tidy a dataset like this, we need to pivot the offending columns into a new pair of variables, it makes "wide" data longer.

```{r}
#template:

#pivot_longer(dataset,
# a set of columns = "The set of columns whose names are values, not variables", 
# names_to = "The name of the variable to move the column names to", 
# values_to = "The name of the variable to move the column values to")
```

```{r}
pivot_longer(table1, c(`1999`,`2000`), names_to="year", values_to="population")
```

### 2. pivot_wider()

*pivot_wider()* is the opposite of *pivot_longer()*. We use it when an observation is scattered across multiple rows. *pivot_longer()* makes wide tables narrower and longer, *pivot_wider()* makes long tables shorter and wider.

```{r}
table2<- tibble(
  `country` = c("Afghanistan", "Afghanistan", "Afghanistan", "Afghanistan", "Brazil", "Brazil", "Brazil", "Brazil","China", "China","China", "China"),
   `year` = c(1999,1999,2000,2000,1999,1999,2000,2000,1999,1999,2000,2000),
   `type` = c("case", "population","case", "population","case", "population","case", "population","case", "population","case", "population"),
   `count` = c(745, 19987071, 2666, 20595360, 37737, 172006362, 80488, 174504898, 212258, 1272915272, 213766, 1280428583)
)
table2
```

```{r}
# template: 

# pivot_wider(dataset, 
#             names_from = "the column to take variable names from", 
#             values_from = "the column to take values from")
```

```{r}

table2 %>%
pivot_wider(names_from = "type", values_from = "count") 

```


### Separate 
*separate()* pulls apart one column into multiple columns
```{r, echo=FALSE}
table3<-tibble(
    `country` = c("Afghanistan", "Afghanistan", "Brazil", "Brazil","China", "China"),
     `year` = c(1999,2000,1999,2000,1999,2000),
    `rate`= c("745/19987071", "2666 / 20595360", "37737 / 172006362", "80488 / 174504898", "212258/1272915272", "213766/1280428583")
)
table3
```


```{r}
separate(table3, rate, into=c('case','population'), sep="/")

```


## Functions in stringr package

All functions that work on strings in the stringr package start with *str_*

1. str_c(): to combine two or more strings

```{r}
str_c("x", "y")
```

2. str_sub(): extract parts of a string. 
str_sub() takes start and end arguments which give the position of the substring
```{r}
x <- c("apple", "eggplant", "banana")

str_sub(x,1,3)
str_sub(x[2],1,3)

#change the first letter to upper case
str_sub(x,1,1)<-str_to_upper(str_sub(x,1,1))
x

```


3. str_view(): matching pattern with regular expressions

Basic matches
```{r}
str_view(x,"an")

# start with ^, end up with $
str_view(x, "^a")
str_view(x, "a$")


```

The next step up in complexity is ., which matches any character (except a newline):
```{r}
str_view(x, ".a.")

```


- \d: matches any digit.
- \s: matches any whitespace (e.g. space, tab, newline).
- [abc]: matches a, b, or c.
- [^abc]: matches anything except a, b, or c.

```{r}
# Look for a literal character that normally has special meaning in a regex

str_view(c("abc", "a.c", "a*c", "a c"), "a[ ]")
```


4. str_detect():to determine if a character vector matches a pattern, it returns a logical vector
```{r}
x <- c("apple", "banana", "pear")
str_detect(x, "e")

stringr::words

stringr::words[str_length(stringr::words)==3]

# What proportion of common words end with a vowel?
mean(str_detect(stringr::words, "[aeiou]$"))

```



5. str_count():to tell how many matches there are in a string:
```{r}
x <- c("apple", "banana", "pear")
str_count(x, "a")


# On average, how many vowels per word?
mean(str_count(stringr::words, "[aeiou]"))
```

6. str_extract(): to extract the actual text of a match string

```{r}
colours <- c(" red ", " orange ", " yellow ", " green ", " blue ", " purple ")
colour_match <- str_c(colours, collapse = "|")  

has_colour <- str_subset(sentences, colour_match)

has_colour

matches <- str_extract(has_colour, colour_match)
matches

```


7. str_replace() and str_replace_all() allow you to replace matches with new strings

```{r}
x <- c("apple", "pear", "banana")
str_replace(x, "[aeiou]", "-")

str_replace_all(x, "[aeiou]", "-")


x <- c("1 house", "2 cars", "3 people")
str_replace_all(x, c("1" = "one", "2" = "two", "3" = "three"))
```


Your Turn >>>>>>>>>>>>>>>>>>>>>


1. Isolate the last letter of every name and create a logical variable that displays whether the last letter is one of "a", "e", "i", "o", "u", or "y".  
2.calculate the total number of names end in a vowel (by `year` and `sex`)   
3. display the results as a line plot, separate by sex.

```{r}
summary(babynames$name)

str_sub(babynames$name, 1,1)

babynames %>%
  mutate(vowel=str_detect(name, "[aeiouy]$")) %>%
  group_by(year, sex) %>%
  summarise(total=sum(vowel)) %>%
  ggplot(aes(x=year, y=total)) + geom_line(aes(col=sex))

```


