---
title: \bf\textsc{\large QTM220 LAB1 Fitting simple linear regression model} 
date: \textsc{\today}
documentclass: article
fontsize: 10pt 
geometry: margin=0.7in 
output: pdf_document
header-includes:
  - \usepackage{utopia}
  - \usepackage{enumerate}
  - \usepackage[onehalfspacing]{setspace}
  - \usepackage{hyperref}
  - \usepackage{enumitem}
  - \newcommand{\benum}{\begin{enumerate}}
  - \newcommand{\eenum}{\end{enumerate}}
  - \newcommand{\bitem}{\begin{itemize}}
  - \newcommand{\eitem}{\end{itemize}}
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, prompt=FALSE, message = FALSE,comment=NA )
```

\large In this Lab example, we will fit a simple linear regression model. After fitting the model, we would obtain the parameter estimates $\hat\beta$ and report the result. In addition to the parameter estimates, fitted values ($\hat{y}$) and residuals would be obtained and saved in the data frame. We will also plot fitted line and residuals. The data frame \emph{wage1} in the \emph{wooldrige} package is used. We are interested in the following model
\begin{align*}
wage=\beta_{0}+\beta_{1}educ+u
\end{align*}
where $wage$ is average hourly earnings and $educ$ is years of education (1976 data).

\normalsize 

\section{0. Remove variables}
```{r}
rm(list=ls(all=TRUE))
```

\section{1. Library relevant packages}
```{r}
if(!require(wooldridge)){
    install.packages("wooldridge")
    library(wooldridge)
}

if(!require(ggplot2)){
    install.packages("ggplot2")
    library(ggplot2)
}
```


\section{2. Load data 'wage1' in the wooldridge package}
```{r}
data("wage1")

# How many observations?
nrow(wage1)

# What are the variables in the data set?
names(wage1)

# Summary of the variables?
summary(wage1)


# Summary of a specific variable? Here we are interested in wage and education.
summary(wage1[,c('wage','educ')]) 
df_wage=wage1[,c('wage','educ')]
```

\section{3. Scatter plot}

```{r}
gout_scatter<-ggplot(df_wage, aes(x=educ, y=wage))+
  geom_point(size=1,color='lightblue3')+
  xlab("education")+ylab("wage")+
  scale_x_continuous(limits=c(0, 20))+
  ggtitle("wage vs years of education")+
  theme(plot.title = element_text(hjust = 0.5,size=13),
        panel.background = element_rect(fill = "white", colour = NA),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.title=element_text(size=13),
        axis.line = element_line(colour = "grey92", size = 1, linetype = "solid")
  )
gout_scatter
```


\section{4. Fit OLS}
```{r}
fit<-lm(wage~educ, data=df_wage)
summary(fit)
```

\section{5. Reporting Result}
\begin{align*}
\widehat{wage}&=-0.905+0.541educ\\
&n=526
\end{align*}
Later $R^{2}$ and standard error of the estimates would be added to the reported result.

\section{6. Interpret the result}

\section{7. Obtain fitted values (fitted line) and residuals}
The fitted line is
\begin{align*}
\widehat{wage}&=-0.905+0.541educ
\end{align*}
and the fitted value exists for each observation. For each observation, using the years of education, $\widehat{wage}$ can be computed. To compute this in R, let's first find out how to extract coefficient from the fit that we have obtained. Useful function would be $coef()$.

```{r}
coef(fit)

fit$coefficients
```

The first location is the intercept parameter estimate and the second location is the slope parameter estimate. The fitted value for each observation can be computed as below which would create a $n\times 1$ vector. I have saved this fitted values vector in the data frame $df\_wage$ using the variable name $wage\_hat\_manual$. Check with the View() function.

```{r}
df_wage$wage_hat_manual=coef(fit)[1]+coef(fit)[2]*df_wage$educ
```


What is the fitted wage when the number of years of education is 9?

```{r}
coef(fit)[1]+coef(fit)[2]*9
fit$coefficients[1]+fit$coefficients[2]*9

coef_noname<-unname(coef(fit))
coef_noname[1]+coef_noname[2]*9

predict(fit,newdata=data.frame('educ'=9))
```


What about the residual? Note that 
\begin{align*}
\hat{u}=wage-\widehat{wage}
\end{align*}
Hence it can be computed as below. I have saved the residual in $df\_wage$ data frame using the variable name $resid\_manual$. Check with the View() function.

```{r}
df_wage$resid_manual=df_wage$wage-df_wage$wage_hat_manual
```

What is the residual when the number of years of education is 9 and the hourly wage is $4 ?
```{r}
4-predict(fit,newdata=data.frame('educ'=9))
```

The fitted values and the residuals can be computed using the built-in command.

```{r}
df_wage$resid=resid(fit)
df_wage$wage_hat=fit$fitted.values
```



\section{8. Plot fitted line}
Previously we have plotted data using scatter plot. On top of the scatter plot above, let's plot the fitted line. This can be done by the fitted values $wage\_hat$ that we have saved in the $df\_wage$ data frame.

```{r}
gout_scatter_fitted<-ggplot(aes(x=educ,y=wage),data=df_wage)+
                    geom_point(size=1, color='lightblue3')+
                    geom_line(size=0.8, aes(x=educ,y=wage_hat),color='navy')+
                    xlab("education")+ylab("wage")+
                    scale_x_continuous(limits=c(0, 20))+
                    ggtitle("wage vs years of education")+
                    theme(plot.title = element_text(hjust = 0.5,size=13),
                    panel.background = element_rect(fill = "white", colour = NA),
                    panel.grid.major = element_blank(),
                    panel.grid.minor = element_blank(),
                    axis.title=element_text(size=13),
                    axis.line = element_line(colour = "grey92", size = 1, linetype = "solid")
  )
  
gout_scatter_fitted
ggsave(gout_scatter_fitted,file="sactter_fitted.eps",device='eps')
```

The same plot can be plotted using the $geom\_smooth()$ function with method \emph{lm}.
```{r}

gout_scatter_fitted2<-ggplot(aes(x=educ,y=wage),data=df_wage)+
                    geom_point(size=1, color='lightblue3')+
                    geom_smooth(method='lm', se=FALSE, color='magenta')+
                    xlab("education")+ylab("wage")+
                    scale_x_continuous(limits=c(0, 20))+
                    ggtitle("wage vs years of education")+
                    theme(plot.title = element_text(hjust = 0.5,size=13),
                    panel.background = element_rect(fill = "white", colour = NA),
                    panel.grid.major = element_blank(),
                    panel.grid.minor = element_blank(),
                    axis.title=element_text(size=13),
                    axis.line = element_line(colour = "grey92", size = 1, linetype = "solid")
  )
  
gout_scatter_fitted2

 
```

\section{9. Residual plot}
In this section, let's create a scatter plot of residuals. After plotting think about the following two questions.

\benum

\item How does the variation of wage levels change as years of education increase? 
\item Do you think  $u\vert educ$ is normally distributed?

\eenum

```{r}
gout_resid_fitted<-ggplot(aes(x=educ,y=resid),data=df_wage)+geom_point(size=1, color="navy")+
                  geom_hline(yintercept=0, linetype="dashed", color='lightblue3')+
                    xlab("education")+ylab("residual")+
                    scale_x_continuous(limits=c(0, 20))+
                    ggtitle("residual scatter plot")+
                    theme(plot.title = element_text(hjust = 0.5,size=13),
                    panel.background = element_rect(fill = "white", colour = NA),
                    panel.grid.major = element_blank(),
                    panel.grid.minor = element_blank(),
                    axis.title=element_text(size=13),
                    axis.line = element_line(colour = "grey92", size = 1, linetype = "solid")
)
gout_resid_fitted

```



\section{10. Extra example if time permits}
Using the data frame \emph{'campus'}, fit the following simple linear regression model,
\begin{align*}
crime=\beta_{0}+\beta_{1}enroll+u,
\end{align*}
where $crime$ is the total campus crimes and $enroll$ is the total enrollment in 1992.
Interpret your result. Obtain the fitted values and residuals and plot similar to the previous example. Please comment.

```{r}
rm(list=ls(all=TRUE))

data('campus')
df_campus=campus[,c('crime','enroll')]
summary(df_campus)

fit_crime<-lm(crime~enroll,data=df_campus)
summary(fit_crime)

df_campus$fitted <- fit_crime$fitted.values
df_campus$resid <- resid(fit_crime)

gout1 <- ggplot(aes(x=enroll, y=crime), data=df_campus) + geom_point(color='lightblue') + 
  geom_line(aes(x=enroll, y=fitted))
gout1

gout2<-ggplot(aes(x=enroll, y=resid), data=df_campus)
```
