---
title: "Working with a Dataset"
author: "Jin Kim"
output: html_document
---
# *Note that BOTH Homework 2 (Due WED, 9/23) AND Quiz 4 (THU, 9/24) will be based on this material.*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Importing data and saving data
```{r}
#Save BOTH the csv file and this rmd file in the same folder (preferably your WORKING DIRECTORY)
gss<-read.csv("gss.csv")  # import dataset in R and store it in gss
dim(gss) #This dataset contains 53474 rows and 14 variables

set.seed(1234) #generating a random number sequence
gss100<-gss[sample(nrow(gss),100),] #store 100 observations into an object called gss100
dim(gss100) 

#check your working directory
getwd() #be sure to check your working directory!

# save a new dataset in your working directory
# write.csv(dataset, "name.csv")
# or... write.csv(dataset,"working directory/givendataname.datatype") 
write.csv(gss100, "smallgss.csv")
```

## 2. Observing data 
```{r}
names(gss100) #get variable names
dim(gss100) # the number of observations and variables
summary(gss100)
head(gss100)  # print the first six observations
sum(is.na(gss$age)) #print a total number of missing values of age in the original dataset
sum(is.na(gss100$age)) #print a total number of missing values of age in the smaller dataset we created
?is.na #learn about is.na function using the built-in help document
```


## 3. Extracting parts
Remember that a dataframe has the similar structure as a matrix [rows, columns], and each variable is a vector
```{r}
#dataset[row, column]
gss100[1,2]

# dataset[rows, columns]
gss100[1:5, c(2,5)] 


#dataset$variable, the whole column
gss100[,7]
gss100$marital

# dataset$variable[rows]
gss100$marital[1:10]
```

## 4. Basic plots: 

See R reference card: 

- one variable: histogram
  
- one categorical: barplot(categorical)

- two variables: scatterplot(two quan), side-by-side boxplot(quan vs cate)

```{r}
# one quantitative variable: 
hist(gss100$age)
mean(gss100$age)
sd(gss100$age)

# one categorical variable:
# barplot(frequency table)
table(gss100$marital)
barplot(table(gss100$marital))

# one numerical vs one categorical
boxplot(gss100$age~gss100$sex)

# plot(x,y) will generate a scatter plot for two numerical variables
```

## 5. Vectors
Almost everything in R is a vector.  All elements(values) of a vector must be of the same type.
```{r}
# str() can tell us 
str(gss100)
```


There are four types of atomic vectors in R.
#### 1) Numeric vectors (integer and Double)
A numeric vector is a list of numbers. Integer and double vectors are known collectively as numeric vectors. (Doubles represent floating point numbers that can not always be precisely represented with a fixed amount of memory.)

The *c()* function is used to collect things together into a vector. The output is 

```{r}
c(1,2,3)
# We can save this to a named object:
x<-c(1:10)  # The : sybmbol used to create a sequence of increasing (or decreasing) values
x
y<-c(30:1)
y

# use the square bracket to extract the element(value)
x[5] 
x[c(1,3,5)]

x[3:7]
x[-c(1,3,5,7,9)]
```


#### seq(), rep(), sample()
There are simple patterned numeric vectors can also be produced using the *seq()* function, the *rep()* function, and the *sample()* function.
```{r}
# seq(start, end, by=1)
seq(0,100,by=5) # the sequence of numbers from 0 to 100 with the increment is 5

# rep(x, times of repetition)
rep(2,5)    # repeat the value 2, 5 times

x<-c(1:5)
rep(x,2)  # repeat x twice

y<-c(10:15)
rep(c(x,y),2)  # repeat x and y, twice

# take a random sample, we can use sample() function. 

# sample(vector, sample size, replace=T/F)

set.seed(1234)
sample(1:10, 5, replace=TRUE) 

# randomly choose 5 numbers from 1 to 30 with replacement
# set.seed() is used to fix the starting point to get a random sample, make sure to get the same sample when the code is executed each time
```


#### Numeric vector arithmetic
Arithmetic can be done on R vectors. Note that the computation is performed *elementwise*.
```{r}
x<-c(1:5)
x*2
x^2
7 %% 2  # the modular operator, the remainder of the division of 7 by 2.
7 %/% 2 

x %% 2 # odd number and even number

y<-c(6:10)

x+y  # the same length
y-x 
```

The operation functions in R can be done on numeric vectors as well.

```{r}
x
sum(x)
mean(x)
length(x)
avg_x<-sum(x)/length(x)
avg_x
```

Example 1:

```{r}
a<-c(1, 5, 6, 8, 9, 15, 20, 19, 10)
# check odd numbers:

sum(a%%2) 

```



### 2) Character vectors
Vectors can be made up of strings of characters instead of numbers. 
```{r}
schoolyear<-c("Freshman","Sophomore","Junior","Senior")

schoolyear

x<-c(2,4, 3,1)
 
# give the names to the numeric vector with the character vector
names(x)<-schoolyear
x
```

#### substr() and paste()

There are two basic operations you might want to perform on character vectors. 

The *substr()* funcion takes subtrings, it takes arguments 
*substr(x, start, stop)*, where x is a vector of characters(strings), and *start* and *stop* say which chacarters to keep. 

The other basis operation is building up strings by concatenation. Use the *paste()* function for this.

*tolower()* and *toupper()* are functions to convert upper case and lower case.

```{r}
substr(schoolyear,1,2)


scoolyear1<-paste(schoolyear,"Year")

paste("The",schoolyear,"Year", sep=" ")

tolower(schoolyear)
```

Example 2:
```{r}
substr(state.name, 1, 1)
# translate all the state names to lower case
tolower(state.name)


```


### 3) Logical vectors

R also supports **logical** vectors, where the values are either *TRUE* or *FALSE*. It is often necessary to test relations when programming. R allows testing of equality and inequality relations using the relational operators: 

1. greater than: $>$, 
2. less than: $<$, 
3. equality: $==$, 
4. inequality: $!=$, $>=$, $<=$, 
5. matching: $ %in% $. 

For example, we can ask 

```{r}
x<-c(1:5)
x>3
3==4

schoolyear %in% "Senior"

```

With a logical data result, *TRUE* has an internal value of 1 and FALSE has an internal value of 0.  Summing a logical will evaluate the total number of *TRUE* results.

```{r}
sum(x>3)
```


The logical operation is very helpful to extract subsets by *subset(logical expression)* 

```{r}
gss100subset<-gss100[gss100$year==2010,]
head(gss100subset)

state.name[substr(state.name,1,1)=="N"]

schoolyear[substr(schoolyear,1,1)=="S"]
```


## Getting help
```{r}
 ?hist()
# then look for specific help by using cmd+f for Mac or ctrl+f for Windows!
```

# *Remember to practice this several times to prepare for Quiz 4 (Thursday, 9/24)! If you have any questions, post it in the discussion board and/or visit office hours!*
